---
title: "Project Report Draft"
author: "Amor Ai, Andrew Noecke, Mykalyster Homberg"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction:

Over the past decade, with the growing influence of social media and heightened popularity of global stars, interest in women’s sports has skyrocketed. However, even with the appearance of high-profile campaigns for equity in sports [1], it has nevertheless been difficult to cultivate an audience and build a market for women’s sports when they receive minimal media coverage compared to their male counterparts [2]. Although major tournaments, such as the WNBA Finals, are strong pulls for fans, the lack of nationally televised games and limited marketing budgets to showcase the players throughout the year has been a barrier for sports fans — especially established NBA fans — to consistently engage with the WNBA even when they’re interested. Furthermore, although we have seen how statistics can fuel sports passion and storytelling, it was only until recently that easy-to-access data and advanced statistics for the WNBA has even been available to the public [3]. Therefore, we seek to not only promote sustained fan engagement and interactions with the WNBA, we hope to also provide better and more accessible statistics of the players as well.

**Our project aims to make the following contributions, which will be displayed in a public facing Shiny App:**

1. **Develop archetypes of current WNBA players based on each player’s tendencies, abilities, and overall statistics**
    + **Drawing comparisons between WNBA players to NBA players using the archetypes**
2. ***Produce a metric, regularized adjusted plus-minus (RAPM), to measure attributes and provide better evaluations of WNBA players**
    + ***Utilize RAPM to create future projections of players**

Ultimately, we believe that developing an attachment to specific WNBA players through comparisons and creating better metrics to measure player performance can boost year-round engagement and bring the WNBA into the spotlight and keep them there for years to come. 

**SHINY APP:**


# Data:

To define player archetypes in the WNBA and compare WNBA players to NBA players, data must be gathered on a seasonal basis. Using Basketball Reference [4], player statistics were gathered dating back to 2018. Relevant variables included:

* Per 100 possession statistics such as:
  + Points, rebounds, assists
  + Turnovers, field goals made, field goals attempted 
  + Advanced statistics designed to demonstrate player tendencies and efficiency such as:
  + PER, Usage Rate, Win Shares
  
* Play by play statistics such as:
  + Type of turnover (bad pass, lost ball)
  + Foul information (fouls drawn, fouls committed)
  
* Shooting tendency data such as:
  + Percentage of shots at the rim, percentage of shots from 3
  + Percentage of shots that were assisted

This data allows both play style and effectiveness to be evaluated and considered when developing player archetypes and subsequently creating player comparisons between WNBA players and NBA players. 



# Methods:

### Developing Archetypes:

Cleaning the 2021 WNBA all stats dataset:

* Players with MPG < 10 and GP < 5 were dropped

* Player sample size was reduced from 183 to 131

* TOT rows were eliminated due to BBallRef miscalculations of TOT rows

* Kept their stats from teams on which they played more games (only 2 players played more than 5G and 10MPG on both teams)

* Player sample size was reduced from 131 to 120

#### PCA:

* Filtering columns before PCA
  + Removed columns from clean wnba_2021_stats dataset that contain percentages and NA’s
  + Removed highly correlated attempted variables (3 pointers attempted, total field goals attempted, etc.)

* PCA results
  + Top 10 dimensions explain 90% of the variability
  + Grabbed top 3 drivers in each of the top 10 PC's and removed repeated variables
  + Checked correlation matrix

* Initial variables from PCA to use in clustering: 
  + fg_tot, shoot_fouls_drawn, per, orb, distance, trb, tov, ortg, oncourt, shoot_fouls_committed, pf_tot, x3p, ft, pts, badpass, drtg, ftr, stl, pf, ftr, on_off, blk, dws

* Try clustering PC’s??

#### CLUSTERING:

* Use variables from PCA to cluster WNBA players (in the 2021 season)


### RAPM Model:
- Metrics Included:





# Results:

# Discussion:

# Acknowledgements:

- Max Horowitz
- Ron Yurko


# References:

[1] https://www.teamheroine.com/blog/the-10-best-womens-sport-campaigns-of-2020

[2] https://www.si.com/sports-illustrated/2021/03/24/womens-sports-gender-study-discrepancy

[3] https://niemanreports.org/articles/covering-womens-sports/

[4] https://www.basketball-reference.com/wnba/years/2022_per_game.html


# R code:

```{r - summary of dataset, echo=FALSE, message = FALSE, warning= FALSE}

library(tidyverse)
library(broom)
library(stringr)
library(factoextra)

wnba_all_stats <- read_csv("/Users/aa/Desktop/CMSAC-22-Basketball/data/wnba_all_stats.csv") #1997-2022

# 2021 season
wnba_2021_stats <- wnba_all_stats %>% 
  select(-c("off_fouls_drawn", "link")) %>%  #remove off_fouls_drawn (all na's) and link column
  mutate(mpg = mp/g) %>% 
  filter(season == "2021") 

#Eliminating players that played less than 10 mpg and 5 games during the 2021 season

wnba_ecdf <- wnba_2021_stats %>% #justification
  ggplot(aes(x = mpg)) + 
  stat_ecdf() +
  geom_rug(alpha = 0.3) +
  theme_bw() +
  labs(x = "mpg",
       y = "Proportion of WNBA players")

wnba_2021_stats <- wnba_2021_stats %>% 
  filter(mpg > 10, g > 5, !(tm == "TOT"))  #Removing TOT players (errors)

#Which players played for more than 1 team during the season
tbl <- table(wnba_2021_stats$player)
#tbl[tbl>1]

#Remove rows for the 2 players that played less games with their 2nd team 

wnba_2021_stats <- wnba_2021_stats %>% 
  slice(-c(122, 115)) #120 players left

head(wnba_2021_stats)

```

```{r - EDA}
#Initial EDA plots

```


```{r - PCA}

#remove columns with percentages/containing na's
wnba_numeric <- wnba_2021_stats %>% 
  select_if(is.numeric) %>% 
  select(-season, -g, -mp, -gs, -mpg,
         -fgpercent, -x3ppercent, -x2ppercent, -ftpercent,
         -fga_2p, -fga_0_3, -fga_3_10, -fga_10_16,
         -fga_16_3p, -fga_3p, -fg_2p, -fg_0_3, -fg_3_10, 
         -fg_10_16, -fg_16_3p, -fg_3p, -fg_2p_asst, -fg_2p_asst,
         -corner_3s_attempt, -corner_3s_percent, -fg_3p_asst) %>% 
  select(-contains("percent"))
  
#LOOK AT CORRELATION MATRIX

library(ggcorrplot)
wnba_cor_matrix <- cor(wnba_numeric)
#ggcorrplot(wnba_cor_matrix)

library(dplyr)
library(tidyr)

corr_table <- cor(wnba_numeric) %>%
  as.data.frame() %>%
  mutate(var1 = rownames(.)) %>%
  gather(var2, value, -var1) %>%
  arrange(desc(value)) %>%
  group_by(value) %>%
  filter(row_number()==1)


#remove highly correlated attempted variables
model_x <- as.matrix(dplyr::select(wnba_numeric, 
                                   -x3pa, -x2pa, -fta, x3par, -fga_tot, 
                                   -x3pa_tot, -x2pa_tot, -fta_tot))

pca_wnba <- prcomp(model_x, center = TRUE, scale = TRUE)


# PC's proportion of variance explained -----------------------------------
## how ‘important’ each Principal Component is

#get_eig(pca_wnba) - top 10 dimensions explain 90% of the variability

dim_plot <- pca_wnba %>%
  tidy(matrix = "eigenvalues") %>%
  ggplot(aes(x = PC, y = percent)) +
  geom_line() + geom_point() +
  geom_hline(yintercept = 1 / ncol(model_x),
             color = "darkred", 
             linetype = "dashed") +
  theme_bw()

scree_plot <- fviz_eig(pca_wnba) #scree plot

library(patchwork)
dim_plot + scree_plot

##Extra biplot
# fviz_pca_biplot(pca_wnba, label = "var",  
#                 alpha.ind = .5, col.var = "darkblue",
#                 alpha.var = .75)


#Table of variables and PC's

pca_var <- pca_wnba %>%
  # Extract variable coordinates
  tidy(matrix = "rotation") %>%
  # Format table form long to wide
  pivot_wider(names_from = "PC", names_prefix = "PC", values_from = "value")%>%
  # Rename column with variable names
  rename(Variable=column) %>% 
  # Take absolute value
  mutate(across(PC1:PC45, ~ abs(.x))) 


# Pick drivers in top 10 PC's (check for low correlations) --------------

pc1 <- pca_var %>%  #fg_tot, shoot_fouls_drawn, per
  select(Variable:PC1) %>% 
  arrange(desc(PC1)) %>% 
  slice(1:10)

var <- get_pca_var(pca_wnba)
a<-fviz_contrib(pca_wnba, "var", axes=1, xtickslab.rt=90) # default angle=45°
plot(a,main = "Variables percentage contribution of first Principal Components")

pc2 <- pca_var %>%  #orb, distance, trb
  select(c(Variable,PC2)) %>% 
  arrange(desc(PC2)) %>% 
  slice(1:10)

pc3 <- pca_var %>%  #tov, ortg, oncourt
  select(c(Variable,PC3)) %>% 
  arrange(desc(PC3)) %>% 
  slice(1:10)

pc4 <- pca_var %>%  #shoot_fouls_committed, pf_tot, x3p
  select(c(Variable,PC4)) %>% 
  arrange(desc(PC4)) %>% 
  slice(1:10)

pc5 <- pca_var %>%  #ft, pts, badpass
  select(c(Variable,PC5)) %>% 
  arrange(desc(PC5)) %>% 
  slice(1:10)

pc6 <- pca_var %>%  #tov, drtg, ftr
  select(c(Variable,PC6)) %>% 
  arrange(desc(PC6)) %>% 
  slice(1:10)

pc7 <- pca_var %>%  #stl, drtg, pf
  select(c(Variable,PC7)) %>% 
  arrange(desc(PC7)) %>% 
  slice(1:10)

pc8 <- pca_var %>% #stl, ftr, oncourt
  select(c(Variable,PC8)) %>% 
  arrange(desc(PC8)) %>% 
  slice(1:10)

pc9 <- pca_var %>%  #on_off, blk, oncourt
  select(c(Variable,PC9)) %>% 
  arrange(desc(PC9)) %>% 
  slice(1:10)

pc10 <- pca_var %>%  #on_off, dws, drtg
  select(c(Variable,PC10)) %>% 
  arrange(desc(PC10)) %>% 
  slice(1:10)

#Checking correlations for non-repeated top 3 variables in each PC

pc_corr <- wnba_2021_stats %>% 
  select(fg_tot, shoot_fouls_drawn, per, 
         orb, distance, trb,
         tov, ortg, oncourt,
         shoot_fouls_committed, pf_tot, x3p,
         ft, pts, badpass,
         drtg, ftr,
         stl, pf,
         ftr,
         on_off, blk,
         dws) %>%
  cor() %>% 
  round(1)

ggcorrplot(pc_corr, 
           hc.order = TRUE,
           type = "lower",
           lab = TRUE)

## INITIAL VARIABLES FROM PCA TO USE IN CLUSTERING:
## fg_tot, shoot_fouls_drawn, per, orb, distance, trb, tov, ortg, oncourt,
## shoot_fouls_committed, pf_tot, x3p, ft, pts, badpass, drtg, ftr,
## stl, pf, ftr, on_off, blk, dws

```








